---
title: "BePhyNE Guide: Fitting your data to BePhyNE"
author: "Sean McHugh"
date: "2022-07-01"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## BePhyNE intro


Hello and thank you for checking out BePhyNE (fka PhyNE). This guide will show how to fit a set of spatial Presence and Absence/PsuedoAbsences with corresponding environmental data (ex WorldClim or envirem) to BePhyNE and recover estimates of unimodal symmatric response curves with 3 parameters: optimum, breadth, and (optionally) tolerance. For each environmental predictor for each species and estimates of the the evolutionary model parameters. Each curve evolves under a separate multivariate brownian motion (mvBM) where the optimum and breadth (and tolerance optionally) evolve given a an ancestral root mean and variance covariance matrix describing the rates of evolution and covariance between each response curve parameter 


NOTE: all environmental predictor response curves are currently set as uncorrelated, if trying to estimate phylogenetic niche in temperature and precipitation, the impact of precipitation and temperature on occurrence would be unlinked (no correlation estimated in species specific ENMs) and also the covariance in response curve parameters are not modeled across predicotrs, only between (such that there is a correlation between precipitation optimum and breadth but not precipitation optimum and temperature optimum)
Covariation in these predictors is very interesting and important next step in improving this models and other similar ones.


#Getting Start: Get the Package

lets install the package and 

```{r, echo=T, message=F,warnings=F,results= "hide"}
library(devtools)
devtools::install_github("sean-mchugh/BePhyNE")
library(BePhyNE)

#other packages
library(treeplyr)
require(readr)
require(tibble)
require(MCMCpack)
require(devtools)
require(coda)
require(devtools)
require(coda)
require(ape)
require(truncdist)
require(geiger)
require(phytools)
require(tidyr)
require(ratematrix)
require(mvMORPH)
require(dplyr)
require(mvtnorm)
require(MultiRNG)
require(Rphylopars)
require(Rcpp)
require(doParallel)
require(corpcor)
require(Matrix)
require(treeplyr)
#require(bayou)
require(crayon)
require(shape)
require(scales)
require(phytools)
require(robustbase)


```

## Getting started: Data Format

To run BePhyNE you need a dataframe loaded in from a file format such as a ".csv" (can be loaded in from an excel spreadsheet saved as a.csv file usually) with a specific column order start at leftmost: 
-   latitude in degrees
-   longitutde in degree
-   species name (this must be the same exact name as the corresponding tip label in your phylogeny), 
-   P/A value (either a 1 for presence or 0 for absence), 
-   Environmental predictors (one per row)

This on the back end gets formatted as a list of lists, were each list includes the species name, the vector of P/A, and vectors for the corresponding environmental predictor values.

Lets in load in the example file which contains Presence absence data for 98 species of Plethodontid salamanders from Eastern North America (ENA_Pleth in a lot of the code) we have two macroclimate predictors from worldclim: average annual precipitation (mm) and average annual temperature (degrees Celsisus)

```{r}
data(ENA_Pleth_PA)

#loading in the data had an issue needed to convert rows back to numeric (except species of course)
ENA_Pleth_PA[,c(1:2,4:6)]<-do.call(cbind,lapply(X = ENA_Pleth_PA[,c(1:2,4:6)], FUN=as.numeric))

#load tree
data(ENA_Pleth_Tree)

tree<-ENA_Pleth_Tree
```

Cool, lets look at some of the details of our dataset, first printing some of the first rows to get a sense of the schema, then checking the number of datapoints by checking how many rows we have. Then the number of unique species. 

We can then extract a vector of unique genus names in the dataset using some janky code, we will just use this to plot the data in environmental and geographic space.
```{r, echo=FALSE, results='asis'}
#knitr::kable(head(ENA_Pleth_PA, 10))
```


```{r}
# number of rows
nrow(ENA_Pleth_PA)

#number of unique species
species=as.list(unique(ENA_Pleth_PA$species))

names(species)<-species

genera<-unique(unlist(lapply(tree$tip.label, function(x) scan(text=x, sep="_", what="", quiet=TRUE)[1])))
```


## Checking the Data

We can plot our data both spatially and over climate space chose a species and plot (we are not going to try plotting all 82 at once)
```{r, fig.show='hold'}

#class(ENA_Pleth_PA)
sp<-species$Aneides_aeneus

#plot occurrence data for species in Envrionmental space ("E-space")
plot(c(min(ENA_Pleth_PA$X1),max(ENA_Pleth_PA$X1)),
     c(max(ENA_Pleth_PA$X2),min(ENA_Pleth_PA$X2)), xlab="precip",ylab="temp",main=paste(sp, "E-space"), col=4)

points(ENA_Pleth_PA[ ( ENA_Pleth_PA$species==sp & ENA_Pleth_PA$y==0 ) ,5:6], col=1)
    
points(ENA_Pleth_PA[ ( ENA_Pleth_PA$species==sp & ENA_Pleth_PA$y==1 ) ,5:6], col=2)

#plot occurrence data for species in Geographic space space ("G-space")
plot(c(min(ENA_Pleth_PA$Lat),max(ENA_Pleth_PA$Lat)),
     c(max(ENA_Pleth_PA$Lon),min(ENA_Pleth_PA$Lon)), xlab="Lat",ylab="Lon",main=paste(sp, "G-space"), col=4)

points(ENA_Pleth_PA[ ( ENA_Pleth_PA$species==sp & ENA_Pleth_PA$y==0 ) ,1:2], col=1)
    
points(ENA_Pleth_PA[ ( ENA_Pleth_PA$species==sp & ENA_Pleth_PA$y==1 ) ,1:2], col=2)
```

So we have a dataset, now it is time to put it into a form BePhyNE can read, and then start setting up the details for our MCMC run. BePhyNE reads a list of lists of your PA dtaa and occurrences over each species such that for one species (the way ENA_Pleth_PA is converted to data_final makes column order essential) (NOTE: Lat and Lon are removed at this stage) \n
data_final[[1]] \n
       $speciesname \n
       $ vector of y (1's and 0's for presences and absences) \n
       $ vector X1 \n
       $ vector X2 \n
       $ vector Xetx \n



```{r}
#
data_final<- lapply(split(ENA_Pleth_PA[3:6],ENA_Pleth_PA$species), as.list)
#replace vector of species names with a single name
for (sp in 1:length(data_final)) {
  
  data_final[[sp]]$species= data_final[[sp]]$species[[1]]
  
}

# match data_final species order to tree tip order 
data_final=data_final[order(match(unlist(lapply(data_final, function(sp) sp$species)), tree$tip.label))]

```



## Setting up the MCMC:Choose Your Priors

Now that we have our dataset we can start the exciting task of setting our priors, we will inform our parameters with informative/weakly informative priors (entirely uninformative uniform priors can lead to some weird behavior especially for niche breadth and mvBM pars)

For our dataset (and any other using pseudo absence data generated from background points or some other unbalanced sampling that doesnt reflect true prevalence) I would recommend only estimating optimum and breadth from the mvBM model, and leave tolerance to only be estimated from the data

```{r}
#specifiy predictor that needs to be fixed, or set to FALSE, and set center to what you want it fixed to, if for sal precip, er the better likely
center_fixed=F

v=c(.05, .05)
k=lapply(1:length(v), function(x) log(v[[x]]/(1-v[[x]])))


#
#identify the range you expect breadth to be in, if 
max_bd<-1.5

min_bd<-.1

bd <-((log(max_bd)-log(min_bd))/4)^2

#get number of environmental predictors

pred=length(data_final$Stereochilus_marginatus[3:length(data_final[[1]])])

#set means for prior distributions for mvBM rates
par.sd.lnorm.meanlog<-list(c(log(.2),  log(bd)) #pred 1 (center, width)
                           ,c(log(.2), log(bd)) #pred 2
)


#set sd for prior distributions for mvBM rates

par.sd.lnorm.sdlog<-list(c(.1, .1) #pred 1
                         ,c(.1, .1) #pred 2
)


#compile sd and means into single objects

par.sd.lnorm <-lapply(1:pred, function(x) cbind(par.sd.lnorm.meanlog[[x]], par.sd.lnorm.sdlog[[x]]))

par.sd.lnorm <-lapply(1:pred, function(x) cbind(par.sd.lnorm.meanlog[[x]], par.sd.lnorm.sdlog[[x]]))



#set pars for prior distributions of mvBM root means

par.mu.norm<-list( matrix(c(  0,  .5, #center(norm)
                       log(0.3), .2), #width (lnorm)
                        nrow = 2, ncol = 2, byrow = TRUE), #pred 1 root prior (center mean, center sd, width mean, width sd)
                  matrix(c(0, .5,
                    log(0.3), .2),
                    nrow = 2, ncol = 2, byrow = TRUE) #pred 2 root prior (center mean, center sd, width mean, width sd)
)


```


Since tolerance are not part of the mvBM in the analysis and only estimated from the glm we will put priors on each tip. with the mean being a max likelihood estimate for tolerance Note what we are doing here is a little janky/circular as we are using max likelihood GLM(NOTE) estimates as the means, this is just a weak prior to place it somewhere between .05 and .99. You could use a flat prior but it will result in BePhyNE taking much longer to mix.  


we do not believe this actually reflect prevalence, but instead error in sampling of the occurrences (broad scale gbif data) and generated absence data (using target group absences filtered to an equal number within the species IUCN expert range map and double the number of presences outside the expert range), see the pre print for details


(NOTE) Generalized linear model: just the function we are fitting for each individual species response curve in PhyNE (logit(y) = B0 + B1 * X1 + B2 * X2)

```{r}
#warning the GLM from this simple one predictor model is going to be messy (lot of "Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred") we dont expect the estimate to be accurate we just are trying to get a rough guess of where tolerance may be, and starting values for our individual species response curves

GLM_only_ml<-suppressWarnings(MLglmStartpars(species_data = data_final,tree = tree, height = NULL))

heights_glm<- lapply(GLM_only_ml$start_pars_bt, function(pred) pred[,3])
```

Now we have all the pieces needed to construct our Prior functions, these are what will be passed through BePhyNE's MCMC to calculate the prior likelihood our parameters 

```{r}

Prior_scale <-lapply(1:pred, function(x) makePrior_ENE(r = 2, p=1, #p is the number of regimes, currently BePhyNE is only capable of modeling one, but stay tuned!
                                                       den.mu = "norm",  
                                                       heights_by_sp = heights_glm[[x]], 
                                                       par.mu = par.mu.norm[[x]], 
                                                       den.sd = "lnorm" , 
                                                       par.sd = par.sd.lnorm[[x]], 
                                                       plot=F))
```




## Setting up the MCMC:Choose Your Starting Values

To get start values we simulate parameters of our model, including individual species response curves and mvBM pars, however because we only can set what the mvBM pars are and the tip responses are subsequently simulated, we dont have a lot of control over what the tip values will be and some potential starting values could be so far off they 

```{r}
# Simulate mvBM and starting individual species responses
repeat{

  startPars_scaled <- priorSim_pars(Prior_scale, tree, dist="norm",hard_coded_heights = heights_glm) 
  
  #break when no tip has a -inf likelihood
  
  if(length(findBadStart(res=startPars_scaled$sim_dat$sim_dat_bt, pa_data=data_final, plot=F))==0){break}


}


findBadStart(res=startPars_scaled$sim_dat$sim_dat_bt, pa_data=data_final, plot=F)


```


## Setting up the MCMC:Choose Your Tuning Parameters

Next you need to set your tuning parameters, how often do we want to be making proposals on any specific parameter, we set these through trial and error on our scaled predictors, each tip response par and mvBM par has its own tuning par in this setup, each response curve proposal type (center_lide, center_mult, width_slide, etc) has its own tuning par 


```{r}

#set base tuning pars for each parameter type
    center_slide=.18
    center_mult=.12
    width_slide=.15
    width_mult=.2
    height_slide=.5
    height_mult=.3

    tuning<-  list(
      niche_prop= lapply(1:pred, function(pred) list(slide=tibble(center=sample(center_slide, length(tree$tip.label), replace=T),   
                                                                  width= sample(width_slide, length(tree$tip.label), replace=T), 
                                                                  height= sample(height_slide, length(tree$tip.label), replace=T) ),
                                                      multi=tibble(center=sample(center_mult , length(tree$tip.label), replace=T),   
                                                                   width= sample(width_mult , length(tree$tip.label), replace=T), 
                                                                   height= sample(height_mult, length(tree$tip.label), replace=T)  )
      )),
      w_mu =  lapply(1:pred, function(pred) list(slide=c(.6,.6),
                                                  multi=c(.6,.6)))
      ,
      w_sd =  lapply(1:pred, function(pred)  list(slide=c(.15,.12),
                                                   multi=c(.15,.12))
      ),
      v_cor       = lapply(1:pred, function(pred) 100)
    )




```





##Run Your MCMC (Fingers Crossed)
```{r}
#  
sets_full<-suppressWarnings(separate.data(data_final,ratio = .5))
  sparse_sp=F
  iterations=1000
  trim_freq= 10
  burnin=0
  chain_end=(iterations-(burnin))/trim_freq
  plot=F
  plot_freq=iterations/5
  v=c(.05, .05)
  k=lapply(1:length(v), function(x) log(v[[x]]/(1-v[[x]])))

  moves_wieghts=c("height" =2,
                  "center" =3,
                  "width"  =3,
                  "theta"  =1,
                  "R_corr" =1,
                  "R_sd"   =1)

  move_prob=c("height"  = moves_wieghts[[1]]/sum(moves_wieghts),
              "center" = moves_wieghts[[2]]/sum(moves_wieghts),
              "width"  = moves_wieghts[[3]]/sum(moves_wieghts),
              "theta"  = moves_wieghts[[4]]/sum(moves_wieghts),
              "R_corr" = moves_wieghts[[5]]/sum(moves_wieghts),
              "R_sd"   = moves_wieghts[[6]]/sum(moves_wieghts))
```


```{r,message=FALSE, warning=F, results= "hide" }
  results<- metro_haste_full_MV(      R_corr_start   = startPars_scaled$R$R_cor
                                    , R_sd_start   = startPars_scaled$R$R_sd
                                    , A_start      = startPars_scaled$A$A_bt
                                    , Prior= Prior_scale
                                    , tree = tree
                                    , tibble_data = startPars_scaled$sim_dat$sim_td_bt
                                    , pa_data=sets_full$training
                                    , iterations=iterations
                                    , burnin=burnin
                                    , move_prob=move_prob
                                    , n=2
                                    , print.i.freq=1000
                                    , print.ac.freq=100
                                    , printing=TRUE
                                    , trim=T
                                    , trim_freq=trim_freq
                                    , H_fixed=F
                                    , tuning=tuning
                                    , center_fixed=center_fixed
                                    , write_file=F
                                    , IDlen=5
                                    , dir=NA
                                    , outname=NA
                                    , prior_only= F
                                    , glm_only  = F
                                    , plot=F
                                    , plot_freq=iterations/5
                                    , plot_file= NA
                                    , True_pars = NA
                                    , k=k
  )
```



To those trying to use their own data, if it didn't break down partway through you probably have a working run! Lets trying to visualize our results

First lets take our colossal list of lists of mcmc chains, lets load those as a list of mcmc objects from the package coda, these are easier to plot (not unless tolerances are provided you will recieve errors saying subscript is out of bound, this is fine and just fills the list positions where those chains would go with the error messages)

```{r, message=FALSE, warning=F,results= "hide"}

mcmc_lt<-BePhyNE_out2coda_mcmc(results, tree)

#lets examine our MCMC chains a litttle 
#note pars for each predictor are lumped together and you can call them through subsetting your list

#Ex:
#subset out the mcmc optimum for environmental predictor 1 by specifiying list 1 with $pred_1 to get predictor 2 you would do  $pred_# 2 and so on, same for the specific response curve parameters too 

#note we didnt estimate tolerance in the mvBM so the chains for those pars are obviously missing

head(mcmc_lt$tip_curves$pred_1$optimum)
head(mcmc_lt$tip_curves$pred_1$breadth)
head(mcmc_lt$tip_curves$pred_1$tolerance)

head(mcmc_lt$A$pred_1$optimum)
head(mcmc_lt$A$pred_1$breadth)
head(mcmc_lt$A$pred_1$tolerance)
head(mcmc_lt$A$pred_2$optimum)
head(mcmc_lt$A$pred_2$breadth)
head(mcmc_lt$A$pred_2$tolerance)

head(mcmc_lt$R_sd$pred_1$optimum)
head(mcmc_lt$R_sd$pred_1$breadth)
head(mcmc_lt$R_sd$pred_1$tolerance)
head(mcmc_lt$R_sd$pred_2$optimum)
head(mcmc_lt$R_sd$pred_2$breadth)
head(mcmc_lt$R_sd$pred_2$tolerance)

head(mcmc_lt$R_cor$pred_1$opt.bre)
head(mcmc_lt$R_cor$pred_1$opt.tol)
head(mcmc_lt$R_cor$pred_1$br_tol )
head(mcmc_lt$R_cor$pred_2$opt.bre)
head(mcmc_lt$R_cor$pred_2$opt.tol)
head(mcmc_lt$R_cor$pred_2$br_tol )


````

Now lets start plotting some of our distributions, and getting posterior distribution using HPDInterval (Highest Posterior Density Interval)


```{r, eval=F}

HPDs<-HPD_list(mcmc_lt,tree, prob=0.95)


````


lets plot some MCMC chains, we can either do that directly for specific parameters using traceplot and dense plot, or plotting all of them using PlotBePhyNEchains() fn, which will return a set of files with the chains and/or densities for each parameter either to your work directory or a specified directory, writing these plots to files is usually the best move as there are a lot of parameters (species responses AND mvBM parameters....thats a lot of plotting)

```{r, fig.show='hold'}

traceplot(mcmc_lt$A$pred_1$optimum)

densplot(mcmc_lt$A$pred_1$optimum)

````


```{r, eval=F}

PlotBePhyNEchains(mcmc_lt)

````


> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
